---
title: "Getting started with AVIOR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with AVIOR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # The commands below read/write files, and `assess` needs the riskmetric
  # engine (and, for some metrics, network access), so the chunks are shown
  # for reference rather than executed while the vignette is built.
  eval = FALSE
)
```

AVIOR is a **local-first evidence compiler for R package validation** in
regulated environments (FDA / EMA / NMPA submissions). It takes your dependency
manifest (`renv.lock`) and a validation policy as input and compiles them into
structured, version-controlled validation evidence — an inventory, risk scores,
and decision records that live in your repository. Like `pytest` or `terraform`,
it lives in your repository and your CI rather than on a server.

This vignette walks through the complete V1 workflow: `init` → `scan` →
`assess` → `review` → `test` → `check` → `bundle` → `verify`. Every command
has both an R function (`avior_*()`) and a CLI form (`avior <command>`); this
vignette uses the R functions.

The workflow compiles the version-controlled inputs into an immutable evidence
bundle containing English HTML and DOCX validation reports, a traceability
matrix, environment fingerprints, snapshots, and a SHA-256 manifest. The final
verification step runs without project context, so an auditor can check either
the bundle directory or its transport zip independently.

```{r setup}
library(avior)
```

## 1. Scaffold the policy tree — `avior init`

`avior_init()` creates the `validation/` contract tree in your project root: a
commented `avior.yml` policy skeleton, plus the `decisions/`, `tests/`,
`evidence/` and `.cache/` directories and a `.gitignore` for the cache and
transport zips. It is idempotent — existing files are never overwritten.

```{r init}
avior_init(".")
```

The generated `validation/avior.yml` is meant to be edited by hand. Its default
weights use metadata- and network-tier metrics only, and the `policy.rationale`
field is a deliberate `TODO`: `avior check` fails until your organisation
records *why* it chose this policy.

You can load and inspect the merged policy (your settings over the PRD defaults)
at any time:

```{r config}
cfg <- avior_config_load(".")
cfg$policy$engine
```

## 2. Identify dependencies — `avior scan`

`avior_scan()` parses `renv.lock`, classifies each package (base / recommended /
contributed / custom), detects which packages your source code calls directly
(with `file:line` provenance), applies any `scope` include/exclude overrides,
and records the lockfile's SHA-256 as the drift baseline.

```{r scan}
avior_scan(".")
```

It writes `validation/inventory.yml`, a git-tracked generated artifact. Output
ordering and serialization are deterministic (FR-X-7 / FR-X-8), so re-running
`scan` over an unchanged project produces a byte-identical file.

## 3. Score risk — `avior assess`

`avior_assess()` scores in-scope packages through the engine adapter
(riskmetric by default). Engine metrics are "goodness" values in `[0, 1]`, and
AVIOR aggregates them into a risk score (`risk = 1 - weighted.mean(goodness,
weights)`), maps risk tiers, and maintains an NA-cause-aware score cache under
`validation/.cache/`.

```{r assess}
# Requires the riskmetric engine to be installed:
#   install.packages("riskmetric")
avior_assess(".")
```

By default only offline metadata/network-tier metrics run. Execution-tier
metrics (for example coverage) are expensive, so they only run under `deep`:

```{r assess-deep}
avior_assess(".", deep = TRUE)
```

Results are written to `validation/scores.yml`, including a `run: {deep,
network}` disclosure so the report can state exactly which metrics were
resolved.

## 4. Record decisions — `avior review`

For every in-scope package that lacks a decision record, `avior_review()` writes
a decision stub pre-filled with the current score snapshot, then validates all
decisions for completeness: signature, rationale, version freshness, use
statements on medium/high tiers, targeted tests where required, and AI dual
sign-off.

```{r review}
avior_review(".")
```

Stubs live under `validation/decisions/<package>.yml` and are never overwritten
once created — the git history of that directory *is* your decision trail.

## 5. Run targeted tests — `avior test`

Decision records with `decision: include_with_tests` name the test files that
must provide fresh passing evidence. Put those files under `validation/tests/`.
Each file must contain exactly one package header:

```{r targeted-test-file}
# avior-package: targetpkg

testthat::test_that("the intended-use result is stable", {
  result <- targetpkg::some_function()
  testthat::expect_equal(result, "expected")
})
```

Run all declared targeted tests and write environment-bound results:

```{r targeted-test-run}
test_result <- avior_test(".")
test_result$status
```

`validation/test-results.yml` binds every result to the tested package version,
the dependency-source SHA-256, and the R version/platform. A failing file, an
all-skipped file, or a file with no passing test is not valid evidence.

## 6. Gate CI — `avior check`

`avior_check()` is read-only: it never re-scores. It aggregates lockfile drift
against the inventory baseline, policy validity, decision completeness,
test-result freshness (including the environment binding), and any
`excluded_but_present` packages, and returns a pass/fail result.

```{r check}
result <- avior_check(".")
result$status
```

Run it in CI: the CLI maps a `fail` status to a non-zero exit code, so a broken
evidence chain stops the pipeline.

```{r check-cli}
# In CI, as a shell command:
#   avior check
# or equivalently:
avior::main(c("check"))
```

`avior_init(".", ci = "github")` or `ci = "gitlab"` can scaffold a workflow
that runs this read-only gate. Existing workflow files are never overwritten.

## 7. Compile the evidence bundle — `avior bundle`

Compilation runs the same gate again. By default, a failed gate blocks output;
`force = TRUE` is reserved for exceptional cases and places an explicit failed
integrity disclosure on the report cover.

```{r bundle}
bundle <- avior_bundle(".", zip = TRUE)
bundle$path
bundle$report_files
```

Every new `validation/evidence/bundle-<UTC>/` directory is immutable and
contains `report.html`, `report.docx`, `traceability.csv`, `environment.json`,
`session-info.txt`, `BUNDLE.yml`, the input `snapshot/`, and
`MANIFEST.sha256`. The built-in report renderer has no Quarto or pandoc runtime
dependency. English (`report.language: en`) is complete; the Chinese locale is
an explicit placeholder that fails closed before any partial bundle is written.

## 8. Verify and archive — `avior verify`

Verify the directory and the optional transport zip independently:

```{r verify}
directory_verification <- avior_verify(bundle$path)
zip_verification <- avior_verify(bundle$zip)
stopifnot(
  directory_verification$status == "pass",
  zip_verification$status == "pass"
)
directory_verification$anchor
```

The returned anchor is the SHA-256 of `MANIFEST.sha256`. Record it outside the
bundle in a git commit, QMS archive record, or independent signature. The
manifest proves internal consistency; it is not by itself an identity or trust
root.

## Where to go next

* The command reference: `?avior_init`, `?avior_scan`, `?avior_assess`,
  `?avior_review`, `?avior_test`, `?avior_check`, `?avior_bundle`, and
  `?avior_verify`.
* The file contracts for every generated artifact are specified in the product
  requirements document, `docs/PRD.md` (§6), which is the project's development
  baseline.

> AVIOR produces signature-ready evidence but does not perform electronic
> signatures or controlled archival. Those activities remain in your existing
> QMS, and compliance responsibility always remains with the sponsor.
